---
title: "bird: Bayesian Imputation for Right-censored survival Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bird: Bayesian Imputation for Right-censored survival Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
run_models <- identical(Sys.getenv("BIRD_RUN_VIGNETTE_MODELS"), "true")
```

## Overview

`bird` provides Bayesian multiple imputation for right-censored survival data.

It supports:

- Parametric models (`weibull`, `exponential`, `lognormal`) via Stan
- Nonparametric LDDP imputation
- Grouped analyses
- Multi-model comparisons

Throughout this vignette we use `survival::lung`.

```{r}
library(bird)
library(survival)

lung <- survival::lung
head(lung[, c("time", "status", "sex", "age")])
```

## One-Function Workflow

For most analyses, use `impute()` directly.

```{r eval = run_models}
set.seed(2026)

fit <- impute(
  lung,
  n_imputations = 5,
  distribution = "weibull",
  time_unit = "days",
  mcmc_options = list(iter_warmup = 200, iter_sampling = 200, chains = 2),
  verbose = FALSE
)

print(fit)
```

```{r eval = !run_models}
# To run model-fitting examples in this vignette:
# Sys.setenv(BIRD_RUN_VIGNETTE_MODELS = "true")
```

Time units:

- `time_unit` defaults to `"days"` if omitted.
- `time_unit` controls labels/reporting and default time-horizon summaries.
- Input times are assumed to already be in that unit (no automatic conversion).

`impute()` auto-detects common time/status columns. If names are unusual, specify them:

```{r eval = FALSE}
fit <- impute(
  lung,
  time = "time",
  status = "status",
  n_imputations = 5,
  distribution = "lognormal"
)
```

Switching engines is just argument changes:

```{r eval = FALSE}
fit_np  <- impute(lung, model = "nonparametric", n_imputations = 5)
fit_grp <- impute(lung, groups = "sex", distribution = "weibull", n_imputations = 5)
```

Direct functions are also available: `bayesian_impute()` (parametric) and `bayes_np_impute()` (nonparametric).

## Looking at Results

```{r eval = FALSE}
print(fit)

plot(fit, type = "survival")
plot(fit, type = "survival", n_curves = 20)
plot(fit, type = "completed_dataset_summary")
plot(fit, type = "completed_dataset_summary", dataset_id = 3)
plot(fit, type = "boxplots_comparison")
plot(fit, type = "boxplots_comparison", n_max = 20)
plot(fit, type = "boxplots_comparison", dataset_indices = c(1, 5, 10))
plot(fit, type = "density")
```

Diagnostics:

```{r eval = FALSE}
plot(fit, type = "posterior")
plot(fit, type = "trace")
plot(fit, type = "pairs")
```

`trace`/`pairs` require `bayesplot` (and `posterior`).

## Grouped Analyses

All group-specific objects (`impute(..., groups = ...)`) use the same plotting API.

```{r eval = FALSE}
print(fit_grp)
plot(fit_grp, type = "survival")
plot(fit_grp, type = "completed_dataset_summary")
plot(fit_grp, type = "survival", combine_groups = FALSE)
plot(fit_grp, type = "boxplots_comparison")
```

For `plot(fit_grp, type = "completed_dataset_summary")`:

- 2 groups: full 4-panel view (histogram, density, survival, boxplot)
- >2 groups: simplified view (survival + boxplot)

### 3-Group Example

```{r eval = run_models}
lung3 <- lung
lung3$age_group <- cut(
  lung3$age,
  breaks = c(-Inf, 60, 70, Inf),
  labels = c("<60", "60-70", ">70"),
  right = TRUE
)

fit_grp3 <- impute(
  lung3,
  groups = "age_group",
  distribution = "weibull",
  n_imputations = 5,
  mcmc_options = list(iter_warmup = 200, iter_sampling = 200, chains = 2),
  verbose = FALSE
)
```

```{r eval = FALSE}
print(fit_grp3)
plot(fit_grp3, type = "survival")
plot(fit_grp3, type = "survival", combine_groups = FALSE)
plot(fit_grp3, type = "completed_dataset_summary")
```

## Comparing Multiple Models

```{r eval = run_models}
cmp <- compare_models(
  lung,
  models = c("weibull", "nonparametric"),
  n_imputations = 5,
  mcmc_options = list(iter_warmup = 200, iter_sampling = 200, chains = 2),
  mcmc = list(nburn = 300, nsave = 500, nskip = 1, ndisplay = 100),
  verbose = FALSE
)
```

```{r eval = FALSE}
print(cmp)
plot(cmp, type = "survival")
plot(cmp, type = "completed_dataset_summary")
```

Accepted model values include `"weibull"`, `"exponential"`, `"lognormal"`, and `"nonparametric"` (aliases `"np"`, `"lddp"`).

For `plot(cmp, type = "completed_dataset_summary")`:

- 2 models: full 4-panel view (histogram, density, survival, boxplot)
- >2 models: simplified view (survival + boxplot)

## Working with Completed Datasets

```{r eval = FALSE}
complete(fit, dataset = 1)
all_sets <- complete(fit)
long_fmt <- complete(fit, format = "long")
```

Each completed dataset includes transparency columns (`original_time`, `original_status`, `was_censored`).

For grouped analyses:

```{r eval = FALSE}
combined <- complete(fit_grp, dataset = 1, groups = "combined")
separate <- complete(fit_grp, dataset = 1, groups = "separate")
```

## Exporting

```{r eval = FALSE}
export(fit, "weibull_results", format = "csv")
export(fit, "weibull_results", format = "rds")

export(fit, "clean_results", format = "csv", include_original = FALSE)
export(fit_grp, "groups", format = "csv", groups = "separate")
```

Currently supported export formats are CSV and RDS.

When exporting multiple datasets to RDS, use `combine = FALSE` to write one file per dataset.

## Advanced Usage

Custom priors and MCMC settings can be supplied directly.

```{r eval = FALSE}
my_priors <- get_default_priors("weibull")
my_priors$mu_log_shape <- 0.2
my_priors$sd_log_shape <- 0.4
my_priors$mu_log_scale <- 0.5
my_priors$sd_log_scale <- 0.3

custom_fit <- impute(
  lung,
  time = "time",
  status = "status",
  distribution = "weibull",
  n_imputations = 20,
  priors = my_priors,
  mcmc_options = list(
    chains = 4,
    iter_warmup = 2000,
    iter_sampling = 2000,
    adapt_delta = 0.9,
    max_treedepth = 12
  ),
  verbose = TRUE
)
```

You can do the same for other distributions (`get_default_priors("lognormal")`, etc.).
Nonparametric runs accept `mcmc = list(nburn = ..., nsave = ..., nskip = ...)` in the same way.

## Troubleshooting

- Columns not found: inspect `colnames(data)` and specify `time`/`status` explicitly.
- Stan compilation errors: confirm C++ toolchain setup.
- MCMC diagnostics: inspect `plot(fit, type = "trace")`, then increase iterations if needed.
- Performance: reduce imputations or use a faster distribution (`"exponential"`).
- Reusing draws: use `generate_complete_datasets()` on stored draws instead of refitting.

Quick data sanity check:

```{r eval = FALSE}
stopifnot(
  all(lung$time > 0),
  all(lung$status %in% c(0, 1)),
  !any(is.na(lung$time)),
  !any(is.na(lung$status))
)
```

## Summary

- `impute()` is the default entry point.
- Parametric vs nonparametric vs grouped behavior is controlled by `distribution`, `model`, and `groups`.
- Use `complete()` and `export()` after fitting.
- Plots provide quality checks and diagnostics.
- Custom priors/MCMC settings are available if needed.
